Viscous Transport in Eroding Porous Media

### 2019-08-19 docs/transport/erosion_transport.tex
worked on introduction

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow-induced erosion deteriorates and reshapes solid material over a
range of scales.  At small scales, such as in groundwater flow,
fluid-mechanical erosion significantly alters the geometry, and
ultimately changes the process of transport through the media.  This
paper extends a quadrature method first outlined by Barnett, Wu, and
Veerapaneni~\cite{bar-wu-vee2015} to simulate dense packings of eroding
bodies, and then studies coarse-grained variables to characterize
transport in eroded geometries.

The authors previous work~\cite{qua-moo2018} developed numerical methods
to simulate erosion of multiple bodies suspended in a Stokes flow.  The
numerical methods can be broadly separated into two classes---a
high-order method to solve the Stokes equations and the erosion rate in
a complex geometry, and a time stepping method to evolve the eroding
bodies.  The fluid solver uses a second-kind integral equation that is
discretized with the trapezoid rule and solved iteratively with fast
multipole accelerated GMRES, and then the trapezoid rule is used to
compute the shear stress which is required to find the erosion rate.
Once the erosion rate is computed, the bodies are eroded using a {\thL}
coordinate system.  This allows us to maintain a discretization of the
eroding bodies that is equally spaced in arclength, and is guaranteed to
not tangle.

@@

Since erosion alters the geometry, it also affects coarse-grained
variables.  In this paper, we examine two particular variables that can
be used to understand transport in porous media---the tortuosity and the
anomalous diffusion rates.   The tortuosity quantifies the twisting of
particle trajectories in a medium, and in laminar flow, the tortuosity
is one.  While a porous media often has a large tortuosity, as erosion
increases the gaps between bodies, preferential channels are created,
large numbers of particles become more laminar, and we expect the
tortuosity to decrease.  Anomalous diffusion is often used to describe
transport in a porous media, and it quantifies the dispersion of
particles in the flow.  Again, as the geometry erodes, particles behave
more as a ballistic particle, and we expect less dispersion.  In this
paper, we compute both the tortuosity and the anomalous diffusion rates
of several geometries that form via erosion.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow in many natural occurring geometries involves flows that vary over
many orders of magnitude.  For example, the length scales of a porous
media can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}.  Porous media flow plays an
important role in several environmental and industrial applications.
Therefore, understanding the dynamics of pore-scale flow is central to
upscale the governing fluid equations, characterize
dispersion~\cite{saf1959}, and quantify mixing. However, because of the
large disparity of length and velocity scales, constitutive laws that
link the microscopic scale to macroscopic scale must be
developed~\cite{mil-chr-imh-mcb-ped1998}.  Therefore, to find closure
formulas, accurate simulations at microscopic scales must be performed
and analyzed.  A few examples of closure formulas and coarse-grained
variables include
pressure-saturation-permeability~\cite{mil-chr-imh-mcb-ped1998}, first
passage time distribution~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosity~\cite{hak-com-den2019,
matyka2008tortuosity, dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, and anomalous
diffusion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}.

In many applications, the fluid transport continuously changes the pore
structure.  This is particularly common in geophysical flows where
erosion alters the porous matrix.  In this work, we investigate flow in
eroded porous media.  By introducing erosion, there is a two-way
coupling between the hydrodynamics and the mechanical alteration of the
complex porous geometry.  Transport in porous geometries that have
undergone erosion finds applications in aquifers~\cite{cve-che-wen1996,
sud1986}, solute transport in groundwater flow~\cite{dag1987,
kon-bre1978}.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that can vary over several orders of
magnitude~\cite{}.  Moreover, the channelization results in a geometry
that is largely heterogeneous and anisotropic.  Others have studied
transport in heterogeneous geometries (for example,
see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000, hak-com-den2019,
suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996, leb-ded-dav-bou2007,
knu-car2005, bel-sal-rin1992}). However, to the best of our knowledge, a
detailed study of transport in eroded geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018}.  The incompressible Stokes
equation are solved by first recasting them as a boundary integral
equation (BIE).  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous diffusion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
diffusion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
\end{align}
where $\uu$ is the fluid velocity.  Since there is no diffusion, all
dispersion is a result of the particle spreading induced by the porous
geometry.

In a porous geometry, the length of a particle path is larger than the
distance travelled if the grains were absent.  To quantify the amount of
winding a particle takes, we use the tortuosity that is defined as the
ratio of the particle paths length to the distance between the inlet and
outlet.  In fractured media, the local tortuosity can be over
2.5~\cite{hym-den-hag-kan2019}, and in porous media, depending on the
porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, matyka2008tortuosity}, or greater than
2~\cite{dud-koz-mat2011}.  The tortuosity of the geometry is defined by
averaging over the entire inlet.  Therefore, the geometry's tortuosity
can be used to characterize average particle
motions~\cite{hak-com-den2019}.

Anomalous diffusion  ...
\cite{cus-hu-den1995}

Gap sizes are important for network models ...

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  The fluid
solver uses a second-kind integral equation that is discretized with the
trapezoid rule and solved iteratively with fast multipole accelerated
GMRES, and then the trapezoid rule is used to compute the shear stress
which is required to find the erosion rate.  The integral
equation formulation only requires discretizing the boundary of the
geometry, the complex porous matrix can be easily resolved.  Once the
erosion rate is computed, the bodies are eroded using a
{\thL}~\cite{hou-low-she1994} coordinate system.  This allows us to
maintain a discretization of the eroding bodies that is equally spaced
in arclength, and is guaranteed to not tangle.

@@
